---
title: "Structural Bioinformatics"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## PDB Database Composition Statistics

Download PDB statistics data as CSV file from: http://www.rcsb.org/stats/summary

```{r load, echo=TRUE}
pdbstats <- read.csv("Data Export Summary.csv", row.names = 1)
```

Looking at the table and adding the percent of structres solved by each method...
```{r table, echo=TRUE}
nstr <- sum(pdbstats$Total)
percent <- round((pdbstats$Total / nstr) * 100, 2)
pdbstats$Percent <- percent
library(knitr)
kable(pdbstats)
```

`r percent[1]`% of structures are solved by X-ray crystallography and `r percent[3]`% by electron microscopy.

```{r protein.pct, echo=FALSE}
prot_pct <- round((sum(pdbstats$Proteins) / nstr) * 100, 2)
```

The proportion of protein entries: `r prot_pct`%

## Introduction to Bio3D
```{r}
library(bio3d)
```
```{r}
pdb <- read.pdb("1hsg")
pdb
attributes(pdb)
```
Note: "attributes()" can be used with any r object.

There are 198 amino acid residues and the two non-protein residues are water and the MK1 compound.
The pdb file stores the atom as a dataframe object.

Print a summary of the coordinate data in $xyz
```{r}
pdb$xyz
```
Examine the row and column dimensions
```{r}
dim(pdb$xyz)
```
Print coordinates for the first two atom
```{r}
pdb$xyz[ 1, atom2xyz(1:2) ]
```
Select all C-alpha atoms (return their indices)
```{r}
ca.inds <- atom.select(pdb, "calpha")
ca.inds
```
Make new pdb files here.


## Aligning Multiple Structures

Download some example PDB files(
```{r}
ids <- c("1TND_B","1AGR_A","1TAG_A","1GG2_A","1KJY_A","4G5Q_A")
files <- get.pdb(ids, split = TRUE)
```

Extract and align the chains we are interested in.
```{r}
pdbs <- pdbaln(files, fit = TRUE)
pdbs
```

PCA of these structures
```{r}
# find invariant core
core <- core.find(pdbs)
# superimpose all structures to core
pdbs$xyz = pdbfit(pdbs, core)
# Perform PCA
pc.xray <- pca(pdbs)
```
```{r}
# Calculate RMSD
rd <- rmsd(pdbs)
# Structure-based clustering
hc.rd <- hclust(dist(rd))
grps.rd <- cutree(hc.rd, k=3)
plot(pc.xray, 1:2, col="grey50", bg=grps.rd, pch=21, cex=1)
```